---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: yes
---

```{r setoptions, echo=FALSE}
require(knitr, quietly=TRUE)
opts_chunk$set(echo=TRUE, warning=FALSE, error=FALSE, message=FALSE)
```

```{r libraries, echo=FALSE}
require(dplyr, quietly=TRUE)
require(ggplot2, quietly=TRUE)
require(gridExtra, quietly=TRUE)
require(scales, quietly=TRUE)
require(pander, quietly=TRUE)
```



## Loading and preprocessing the data

Load the data  
```{r dataload}
activity <- read.csv("activity.csv", stringsAsFactors = FALSE)
str(activity)
```

Process/transform the data (if necessary) into a format suitable for your analysis
```{r activitysummary}
activity_summary <- activity %>%
                    group_by(date) %>% 
                    summarise("total"  = sum(steps), 
                              "mean"   = mean(steps), 
                              "median" = median(steps), 
                              "nas"    = sum(is.na(steps)))
head(activity_summary)
```

## What is mean total number of steps taken per day?

Make a histogram of the total number of steps taken each day   
```{r Histogram of Steps per Day, echo=FALSE, fig.width=10}
mean_txt <- paste0("mean = ", round(mean(activity_summary$total, na.rm=TRUE), 2))
median_txt <- paste0("median = ", round(median(activity_summary$total, na.rm=TRUE), 2))

g1 <- ggplot(data=activity_summary, aes(x=total)) +
    geom_histogram(aes(fill = ..count..)) +
    scale_fill_gradient("Count", low = "yellow", high = "red") +
    geom_vline(xintercept=mean(activity_summary$total, na.rm=TRUE), col="green", size=2) +
    annotate("text", x = 17000, y = 7.5, label = mean_txt) +
    annotate("text", x = 17000, y = 7, label = median_txt) +
    guides(fill=FALSE) +
    ggtitle("Number of steps per day") +
    xlab("Total Number of steps taken per day") +
    ylab("Number of days with the same count") +
    theme_bw()

g2 <- ggplot(data=activity_summary, aes(x=as.Date(date), y=total, fill=total)) +
    geom_histogram(stat="identity") +
    geom_hline(yintercept=mean(activity_summary$total, na.rm=TRUE), col="green", size=2) +
    guides(fill=FALSE) +
    ggtitle("Total number steps taken per day") +
    xlab("Date") +
    ylab("Total number of teps") +
    theme_bw()

grid.arrange(g1, g2, ncol=2)
```

Calculate and report the mean and median total number of steps taken per day   
```{r mean and median}
mean(activity_summary$total, na.rm=TRUE)
median(activity_summary$total, na.rm=TRUE)
```


## What is the average daily activity pattern?

Time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r daytime}
activity_daily <- activity %>%
                 group_by(interval) %>%
                 summarise("steps_ave" = mean(steps, na.rm=T))
```

```{r echo=FALSE, echo=FALSE, fig.align='center'}
max_interval <- activity_daily$interval[which.max(activity_daily$steps_ave)]
max_label <- paste0("max step interval = ", max_interval)

g <- ggplot(data=activity_daily, aes(interval, steps_ave)) +
     geom_line(col = "blue", size = 2, alpha = 0.8) +
     geom_vline(xintercept = max_interval, col="red", size = 2, alpha = 0.6) +
     annotate("text", x = 1800, y = 150, label = max_label) +
     ggtitle("Average daily activity") +
     xlab("Time, hours") +
     ylab("Average number of steps") +
     scale_x_continuous(breaks= c(0, 600, 1200, 1800, 2355), labels=c(0, 6, 12, 18, 24)) +
     theme_bw()
    
print(g)
```

The following 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps:

```{r}
activity_daily$interval[which.max(activity_daily$steps_ave)]
```
Which correspnds to 8:35 in the morning.

## Imputing missing values

Total number of missing values in the data set is `r sum(is.na(activity$steps))`. Total number of 5 minute intervals per da is `r 24*60/5`. From activty summary statistics we determien that we are missing data for `r length(which(activity_summary$nas==24*60/5))` days.

For days that are missing values we will use the prevous day as the imputed value except for teh day one where day 2 values will be used:

```{r imputing missing values}
act_imputed <- activity
act_imputed$date <- as.factor(act_imputed$date)
act_list <- split(act_imputed, act_imputed$date)

na_days <- which(activity_summary$nas == 24*60/5)
nmax <- length(act_list)

for(i in na_days) {
    if(i == 1) {
        act_list[[i]]$steps <- act_list[[i+1]]$steps
    } else {
        act_list[[i]]$steps <- act_list[[i-1]]$steps
    }
}

act_imputed <- unsplit(act_list, act_imputed$date)
act_imputed$date <- as.Date(act_imputed$date)

act_imp_sum <- act_imputed %>%
               group_by(date) %>% summarise("total"  = sum(steps))
```

Let's look at the histogram of the daily total number of steps for the new dataset with imputed values:
```{r daily average activity for imputed data set, echo=FALSE, fig.width=10}
imp_mean_txt <- paste0("mean = ", round(mean(act_imp_sum$total), 2))
imp_median_txt <- paste0("median = ", round(median(act_imp_sum$total), 2))

g1 <- ggplot(data=act_imp_sum, aes(x=total)) +
    geom_histogram(aes(fill = ..count..)) +
    scale_fill_gradient("Count", low = "yellow", high = "red") +
    geom_vline(xintercept=mean(act_imp_sum$total, na.rm=TRUE), col="purple", size=2) +
    annotate("text", x = 17000, y = 7.5, label = imp_mean_txt) +
    annotate("text", x = 17000, y = 7, label = imp_median_txt) +
    guides(fill=FALSE) +
    ggtitle("Number of steps per day") +
    xlab("Total Number of steps taken per day") +
    ylab("Number of days with the same count") +
    theme_bw()

g2 <- ggplot(data=act_imp_sum, aes(x=as.Date(date), y=total, fill=total)) +
    geom_histogram(stat="identity") +
    geom_hline(yintercept=mean(act_imp_sum$total, na.rm=TRUE), col="purple", size=2) +
    guides(fill=FALSE) +
    ggtitle("Total number steps taken per day") +
    xlab("Date") +
    ylab("Total number of teps") +
    theme_bw()

grid.arrange(g1, g2, ncol=2)
```

Calculate and report the mean and median total number of steps taken per day  for the data set with imputed values.
```{r imputed data set mean and median}
mean(act_imp_sum$total)
median(act_imp_sum$total)
```
The new values for mean and median are lower for the imputed data set than mean and mediam for the original data set with missing values.

## Are there differences in activity patterns between weekdays and weekends?

```{r weekday factor}
act_imputed$daytype <- "weekday"
act_imputed$daytype[weekdays(act_imputed$date) %in% c("Saturday", "Sunday")] <- "weekend"
act_imputed$daytype <- as.factor(act_imputed$daytype)

act_wkd <- act_imputed %>% 
           group_by(daytype, interval) %>%
           summarize(steps_ave = mean(steps))

str(act_wkd)
```

```{r Weekend and weekday average activity, echo=FALSE, fig.align='center'}
g <- ggplot(act_wkd, aes(interval, steps_ave)) + 
     geom_line(colour = "blue", size = 2, alpha = 0.8) +
     facet_grid(daytype ~ .) +
     ggtitle("Average daily activity") +
     xlab("Time, hours") +
     ylab("Average number of steps") +
     scale_x_continuous(breaks= c(0, 600, 1200, 1800, 2355), labels=c(0, 6, 12, 18, 24)) +
     theme_bw()

print(g)
```

